Species motif participation provides unique information about species risk of extinction

### 2023-06-06 manuscript/JAE_reply.tex
Reply ready for JAE

    \textbf{R:} We thank the Reviewer for taking the time to read the manuscript and for their thoughtful feedback. We have redone our analyses with binomial error distribution as suggested (though with glm rather than glmer as the random effects proved singular). Our main results did not change much with this improved approach, though some of our supplemental results (e.g., those relating to overall network structure) were simplified. We hope that the revised manuscript is now clearer and easier to read and that our statistical approach is more reasonable. Below, we respond to each of the Reviewer's comments (preceded by \textbf{R:}) in turn.

@@

    \textbf{R:} We thank the Reviewer for taking the time to read our manuscript and for the thoughtful feedback. We particularly appreciate the pointers where the structure of our manuscript was unclear an where it was difficult to extract our main aims and results. In the present revision we have substantially reworked the introduction and discussion and reordered the methods and results in order to present our work in a more streamlined and, hopefully, understandable manner. We believe that the manuscript has been greatly improved as a result. Note that, in response to a comment from the other Reviewer, we have also redone our analyses and some results have slightly changed from the previous submission. Below, we respond to each of the Reviewer's comments (preceded by \textbf{R:}) in turn.

@@

                \textbf{R:} A focal species is simply that for which the motif role is currently being calculated ... all consumers are taken as the focal species in turn. Note that as these networks are simulated the species (nodes) in them do not correspond to any particular focal species or trophic guild. The vagueness is intentional as there are no characteristics of, say, invertebrates vs. vertebrates or mammals vs. reptiles incorporated in the networks.

@@

                \textbf{R:} Previous research has shown that omnivory can either be stabilising or destabilising, depending on the precise disturbance, strengths of interactions, etc. Our results, which deal specifically with persistence after bottom-up disturbance, add to that picture of complexity without contradicting earlier findings. Note that we do not say that omnivory is never beneficial - only that in our networks, in response to our focal disturbance, is is generally detrimental. We also note that our results show apparent competition as associated with less persistence when disturbance is weak, which is compatible with the possibility that the Reviewer mentions above (though we do not specifically model prey switching, which seems to be what the Reviewer refers to).  In the revision, we have expanded our discussion on how the strength and direction of motif participation-persistence relationships vary network-to-network (especially in networks with different connectance). We hope that this addition provides some of the nuance the Reviewer missed from the previous revision.

@@

                \textbf{R:} Using the Bayesian network framework, we simply cannot include top-down effects on persistence. However, in a parallel study (started around the same time but by chance published earlier) some of the authors of the present study did simulate both top-down and bottom-up effects in a dynamical model. The dynamical model has its own caveats (in general, dynamical models are sensitive to parameter choice and, due to the time and computational resources required, a full sensitivity analysis could not be performed). However, we note that a dynamical analysis on the same set of networks (albeit with a different disturbance and persistence metric) found substantial consistency in motif-persistence relationships.

                It may also reassure the Reviewer to know that in the initial studies introducing the Bayesian network framework~\citep{Eklof2013}, Bayesian networks captured the majority of the extinctions identified by dynamical models run on the same set of networks. As we use the same modelling framework as in those studies, we can reasonably expect similar results.

                Finally, consider the situation from an ecological perspective. We simulate disturbance to primary producers only (and appreciate that this may not have been clear in the previous revision). While predation certainly does influence persistence, the disturbance we introduce will first act to remove food sources from some herbivores. Some fraction of these herbivores will then go extinct, reducing resources for their predators. Some fraction of these predators will also go extinct due to the reduction in their food. It is only those predators which are affected by declining resources, do not go extinct, and are able to increase predation on their remaining prey which can cause top-down extinctions. While this obviously does occur in nature, logically this is a smaller number of extinctions than the set of herbivore and predator extinctions caused by bottom-up effects. This is also reflected in the similar results from Bayesian networks and dynammical models in the studies cited above - if top-down effects were responsible for the majority of secondary extinctions we would see quite different performance.

                None of this is to discount the reality of top-down extinctions or their interest from a scientific standpoint. However, since the Bayesian network framework does not include top-down effects any comment we might make about how top-down effects might affect our results would necessarily be pure speculation. Instead, we have added a reference to another paper (quoted below) exploring the relationship between motif roles and persistence in a dynamical context. That paper, due to the computational limits inherent in the dynamical approach, uses only a single-species disturbance. It does, however, include top-down as well as bottom-up effects and may provide the information the Reviewer is seeking.

                \begin{quotation}
    There are limitations to the approach presented here.
    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2022Oikos}.
    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.
    Another potentially useful addition to future work would be to include species traits such as body size, which are related to both motif participation~\citep{cirtwill2018feeding} and extinction risk~\citep{Brose2017,Cardillo2005,Purvis2000} in empirical webs.

                \end{quotation}

>>>>

    \textbf{R:} We thank the Reviewer for taking the time to read the manuscript and for their thoughtful feedback. We have redone our analyses with binomial error distribution as suggested (though usually with glm rather than glmer as the random effects proved singular). Our main results did not change much with this improved approach, though some of our supplemental results (e.g., those relating to overall network structure) were simplified. We hope that the revised manuscript is now clearer and easier to read and that our statistical approach is more reasonable. Below, we respond to each of the Reviewer's comments (preceded by \textbf{R:}) in turn.

@@

    \textbf{R:} We thank the Reviewer for taking the time to read our manuscript and for the thoughtful feedback. We particularly appreciate the pointers where the structure of our manuscript was unclear and where it was difficult to extract our main aims and results. In the present revision we have substantially reworked the introduction and discussion and reordered the methods and results in order to present our work in a more streamlined and, hopefully, understandable manner. We believe that the manuscript has been greatly improved as a result. Note that, in response to a comment from the other Reviewer, we have also redone our analyses and some results have slightly changed from the previous submission. Below, we respond to each of the Reviewer's comments (preceded by \textbf{R:}) in turn.

@@

                \textbf{R:} A focal species is simply that for which the motif role is currently being calculated. All consumers are taken as the focal species in turn. Note that as these networks are simulated the species (nodes) in them do not correspond to any particular real species or trophic guilds. The vagueness about species identity is intentional as there are no characteristics of, say, invertebrates vs. vertebrates or mammals vs. reptiles incorporated in the networks.

@@

                \textbf{R:} Previous research has shown that omnivory can either be stabilising or destabilising, depending on the precise disturbance, strengths of interactions, etc. Our results, which deal specifically with persistence after bottom-up disturbance, add to that picture of complexity without contradicting earlier findings. Note that we do not say that omnivory is never beneficial (indeed, it is generally associated with higher persistence in some highly-connected webs, as shown in Fig. 3). However, in our networks overall, in response to our focal disturbance, omnivory is generally detrimental. We also note that our results show apparent competition as associated with less persistence when disturbance is weak, which is compatible with the possibility that the Reviewer mentions above (though we do not specifically model prey switching, which seems to be what the Reviewer refers to, this is included in the ability of a consumer to persist on a subset of its prey).  In the revision, we have expanded our discussion on how the strength and direction of motif participation-persistence relationships vary network-to-network (especially in networks with different connectance). We hope that this addition provides some of the nuance the Reviewer missed from the previous revision.

@@

                \textbf{R:} Using the Bayesian network framework, we simply cannot include top-down effects on persistence. However, in a parallel study (started around the same time but by chance published earlier) some of the authors of the present study did simulate both top-down and bottom-up effects in a dynamical model. Although that study has its own caveats (in general, dynamical models are sensitive to parameter choice and, due to the time and computational resources required, a full sensitivity analysis could not be performed), we refer readers interested in combined top-down and bottom-up effects there. 
                It may also reassure the Reviewer to know that in the initial studies introducing the Bayesian network framework~\citep{Eklof2013}, Bayesian networks captured the majority of the extinctions identified by dynamical models run on the same set of networks. As we use the same modelling framework as in those studies, we can reasonably expect similar results.

                Finally, consider the situation from an ecological perspective. We simulate disturbance to primary producers only (and appreciate that this may not have been clear in the previous revision). While predation certainly does influence persistence, the disturbance we introduce will first act to remove food sources from some herbivores. Some fraction of these herbivores will then go extinct, reducing resources for their predators. Some fraction of these predators will also go extinct due to the reduction in their food. It is only those predators which are affected by declining resources, do not go extinct, and are able to increase predation on their remaining prey which can cause top-down extinctions. While this obviously does occur in nature, logically this is a smaller number of extinctions than the set of herbivore and predator extinctions caused by bottom-up effects. This is also reflected in the similar results from Bayesian networks and dynamical models in the studies cited above - if top-down effects were responsible for the majority of secondary extinctions we would see quite different performance.

                None of this is to discount the reality of top-down extinctions or their interest from a scientific standpoint. However, since the Bayesian network framework does not include top-down effects any comment we might make about how top-down effects might affect our results would necessarily be pure speculation. Instead, we have added a reference to our dynamical paper (quoted below). That paper, due to the computational limits inherent in the dynamical approach, uses only a single-species disturbance and a different estimate of persistence. It does, however, include top-down as well as bottom-up effects and may provide the information the Reviewer is seeking.

                \begin{quotation}
                    There are limitations to the approach presented here.
                    Most importantly, the Bayesian network framework operates on a strict bottom-up principle and therefore cannot include top-down effects in the extinction risk calculations~\citep{Eklof2013}. 
                    Here our focus was to analyze bottom-up affects following disturbances to the basal resources, in which case the Bayesian network approach captures a majority of the secondary extinctions obtained in a fully-fledged dynamical model with much greater efficiency~\citep{Eklof2013}.
                    If top-down effects are important for a particular empirical system, however, our simulations may not be accurate. 
                    For example, although a recent study using a dynamical model to relate motif participation to persistence after a disturbance found similar trends for omnivory, participation in the direct competition motif was not universally positive~\citep{Cirtwill2022Oikos}.
                    In particular, species which frequently appeared in the prey position of the direct competition motif tended to have shorter persistence times after disturbance in that study - possibly due to top-down effects from the pair of predators involved.
                    However, as these two studies applied different types of disturbances, further study would be required to disentangle the effects of top-down and bottom-up effects on species with different motif participation, and how each relates to the type and strength of disturbance.
                    Another potentially useful addition to future work would be to include species traits such as body size, which are related to both motif participation~\citep{cirtwill2018feeding} and extinction risk~\citep{Brose2017,Cardillo2005,Purvis2000} in empirical webs.
                \end{quotation}
